---
title: "Homework 4"
# author: "Type Your Name Here"
# date: today
jupyter: stats
format: pdf
show_solution: true
---


logistic lda




# Week 4: Classification Methods

## Theory Problems

1. Explain logistic regression.

2. Derive the logistic function

$$
P(Y=1|X)=\frac{1}{1+\exp(-\beta_0-\beta^TX)}
$$

3. Explain decision boundaries.

4. Explain LDA assumptions.

5. Compare logistic regression and LDA.

6. Explain Bayes classifier.

7. Explain classification error.

8. Explain ROC curves.

9. Explain precision vs recall.

10. Explain confusion matrix.





# Week 4: Classification, Logistic Regression, LDA, and Trees

## Problem 1: Logistic Regression Model

For binary classification, logistic regression assumes

$$
P(Y=1 \mid X)
=
\frac{1}{1+\exp(-\beta_0-\beta^\top X)}.
$$

1. Explain why this probability is always between 0 and 1.
2. Show that the log-odds are linear in $X$:

   $$
   \log\frac{P(Y=1 \mid X)}{1-P(Y=1 \mid X)}
   =
   \beta_0+\beta^\top X.
   $$

---

## Problem 2: Classification Rule

Suppose a fitted logistic model gives

$$
\hat P(Y=1 \mid X=x)=0.73.
$$

1. Classify the observation using the threshold $0.5$.
2. Explain how the classification changes if the threshold is $0.8$.

---

## Problem 3: LDA Assumptions

LDA assumes that within each class the predictors are multivariate normal with class-specific means and a common covariance matrix.

1. Explain why this leads to a linear decision boundary.
2. Explain when LDA may outperform logistic regression.

---

## Problem 4: Bayes Classifier

1. Explain the idea of the Bayes classifier.
2. Explain why it is theoretically optimal.
3. Explain why it is not directly available in practice.

---

## Problem 5: Trees

1. Explain how a decision tree recursively partitions the predictor space.
2. Explain why the prediction regions are piecewise constant.
3. Explain one advantage and one disadvantage of trees.

---

## Problem 6: Heart Disease Data — Logistic Regression

Use the UCI Heart Disease dataset or another open binary classification dataset.

1. Split the data into training and testing sets.
2. Fit a logistic regression model.
3. Report test accuracy, sensitivity, and specificity.
4. Report the confusion matrix.

```{r}
# Your code here
```

---

## Problem 7: Logistic Regression Interpretation

Using your fitted logistic model:

1. Select one numerical predictor.
2. Interpret its coefficient in terms of odds ratio.
3. Comment on whether the sign matches your intuition.

```{r}
# Your code here
```

---

## Problem 8: LDA Comparison

Using the same train/test split:

1. Fit an LDA classifier.
2. Compare its accuracy with logistic regression.
3. Comment on which method performs better and why that might happen.



---

## Problem 9: Decision Tree

Using the same dataset:

1. Fit a classification tree.
2. Plot the tree.
3. Report test accuracy.
4. Compare with logistic regression and LDA.

```{r}
# Your code here
```

---

## Problem 10: ROC Curve and Thresholding

Using the logistic regression model:

1. Compute predicted probabilities on the test set.
2. Draw the ROC curve.
3. Compute AUC.
4. Explain why accuracy alone may not fully describe classifier performance.

```{r}
# Your code here
```