




# Week 3: Flexible and Nonparametric Models

## Theory Problems

1. Describe the KNN regression algorithm.

2. Explain how $K$ controls model flexibility.

3. Discuss the bias–variance tradeoff in KNN.

4. Explain the curse of dimensionality.

5. Explain why KNN performs poorly in high dimensions.

6. Compare parametric vs nonparametric regression.

7. Explain how kernel smoothing works.

8. Explain the role of distance metrics in KNN.

9. Explain why larger datasets benefit KNN.

10. Discuss advantages and disadvantages of KNN.

---





# Week 3: KNN, Flexible Regression, and Nonparametric Methods

## Problem 1: KNN Regression Estimator

The KNN regression estimator at $x_0$ is

$$
\hat f(x_0)
=
\frac{1}{K}
\sum_{i \in N_K(x_0)} y_i,
$$

where $N_K(x_0)$ is the set of indices of the $K$ nearest training points.

1. Explain how this estimator works.
2. Explain why $K=1$ gives a very flexible fit.
3. Explain why large $K$ gives a smoother fit.

---

## Problem 2: Bias-Variance in KNN

1. Explain how increasing $K$ changes bias.
2. Explain how increasing $K$ changes variance.
3. Explain why there is usually an optimal value of $K$.

---

## Problem 3: Curse of Dimensionality

1. Explain why nearest neighbors become less meaningful in high dimensions.
2. Explain why nonparametric methods often require more data than parametric methods.

---

## Problem 4: Distance Metrics

1. Explain why standardization matters in KNN.
2. Explain how one predictor with a much larger scale can distort the neighborhood structure.

---

## Problem 5: Comparison with Linear Regression

1. Give an example where linear regression is preferable to KNN.
2. Give an example where KNN is preferable to linear regression.

---

## Problem 6: Simulated Nonlinear Regression

Generate data from

$$
y = \sin(2\pi x) + \varepsilon,
\qquad
\varepsilon \sim N(0, 0.2^2).
$$

1. Fit linear regression.
2. Fit polynomial regression of degree 5.
3. Fit KNN regression for several values of $K$.
4. Compare test MSE.

```{r}
set.seed(432)
# Your code here
```

---

## Problem 7: KNN on the Bike Sharing Data

Use the Bike Sharing dataset.

1. Select a response variable related to rental counts.
2. Split the data into training and testing sets.
3. Fit KNN regression for $K=1,3,5,10,20,50$.
4. Plot test MSE versus $K$.
5. Identify the best $K$.

```{r}
# Your code here
```

---

## Problem 8: Effect of Standardization in KNN

Using the same data:

1. Fit KNN without standardization.
2. Fit KNN with standardization.
3. Compare the results and explain the difference.

```{r}
# Your code here
```

---

## Problem 9: KNN versus Linear Regression

Using the same train/test split:

1. Fit a linear regression model.
2. Compare its test MSE with the best KNN model.
3. Explain which model appears more suitable and why.

```{r}
# Your code here
```

---

## Problem 10: Smoothing Interpretation

Use a one-dimensional predictor/response pair from any open dataset.

1. Fit KNN or another smoother.
2. Plot the fitted curve.
3. Explain how the method captures nonlinearity.

```{r}
# Your code here
```