---
title: "Homework 3"
# author: "Type Your Name Here"
# date: today
jupyter: stats
format: pdf
show_solution: true
---



PCR PLS



::: {#exm-}
ISRL 9
:::







::: {#exm-}

## PCR vs PLS

Consider a regression problem with centered data matrix $X \in \mathbb{R}^{n \times p}$ and response $y \in \mathbb{R}^n$.

**a. Bias–Variance Tradeoff**

Discuss how PCR and PLS differ in terms of bias and variance.

- When might PCR have higher bias than PLS?
- When might PLS have higher variance than PCR?


**b. When PCR Fails**

Construct (or describe) a situation where PCR performs way worse than PLS. Explain why PCR may perform poorly in this setting.
<!-- 
**c. When PLS Fails**

Construct (or describe) a situation where PLS performs way worse than PCR. Explain why PLS may perform poorly in this setting. -->

**c. Code examples**

Write Python code to generate the datasets to implement the above two cases. A dataset is a pair $(X,y)$ with both are `numpy.array`.

Ensure there is a performance difference. You can get a bonus if there is at least 1% difference in $R^2$. You may fix random seeds or adjust noise levels to achieve this.

You code should look like this.

```{python}
#| eval: false
import numpy as np

rng = np.random.default_rng(0)

## Write your data generating code here to replace the code below
n, p = 100, 100
X = rng.normal(size=(n, p))
y = rng.normal(size=n)
```

I will test the dataset using the following setting:

1. Split the dataset into trainning and test set with `test_size=0.3`, `random_state=0`.
2. Use `cv=5` and `neg_mean_squared_error` to pick `n_components`. You may set the maximal search cap to be `20` for simplicity.
3. Train the model on the training set, and test the model on the test set.
4. For performance I only look at `r2_score`.


```{python}
#| eval: false
#| echo: false
import numpy as np
from sklearn.model_selection import train_test_split, GridSearchCV
from sklearn.cross_decomposition import PLSRegression
from sklearn.decomposition import PCA
from sklearn.linear_model import LinearRegression
from sklearn.pipeline import Pipeline
from sklearn.preprocessing import StandardScaler
from sklearn.metrics import mean_squared_error, r2_score


def compare_pcr_pls(X, y, max_components=20):

    X_tr, X_te, y_tr, y_te = train_test_split(X, y, test_size=0.3, random_state=0)

    max_components = min(max_components, X_tr.shape[0], X_tr.shape[1])
    param_range = list(range(1, max_components + 1))

    pcr_pipe = Pipeline(
        [
            ("scaler", StandardScaler()),
            ("pca", PCA()),
            ("ols", LinearRegression()),
        ]
    )

    pcr_grid = GridSearchCV(
        pcr_pipe,
        {"pca__n_components": param_range},
        cv=5,
        scoring="neg_mean_squared_error",
    )

    pls_pipe = Pipeline(
        [
            ("scaler", StandardScaler()),
            ("pls", PLSRegression()),
        ]
    )

    pls_grid = GridSearchCV(
        pls_pipe,
        {"pls__n_components": param_range},
        cv=5,
        scoring="neg_mean_squared_error",
    )

    pcr_grid.fit(X_tr, y_tr)
    pls_grid.fit(X_tr, y_tr)

    pcr_mse = mean_squared_error(y_te, pcr_grid.predict(X_te))
    pls_mse = mean_squared_error(y_te, pls_grid.predict(X_te))
    pcr_r2 = r2_score(y_te, pcr_grid.predict(X_te))
    pls_r2 = r2_score(y_te, pls_grid.predict(X_te))

    print(f"PCR test MSE: {pcr_mse}, PCR test R2: {pcr_r2}")
    print(f"PLS test MSE: {pls_mse}, PLS test R2: {pls_r2}")


compare_pcr_pls(X, y)

```


:::

::: {.content-visible when-meta="show_solution"}

::: {.solution}
 **(a) Bias–Variance Tradeoff**

PCR and PLS differ in how they construct components, which leads to different bias–variance behavior.

- **PCR** selects components based only on the variance of $X$, ignoring $y$.
- **PLS** selects components based on their covariance with $y$.

**Bias**

- PCR may have **higher bias** when the directions that are most predictive of $y$ correspond to **low-variance directions in $X$**.  
  In this case, PCR may discard these directions when truncating components.

- PLS tends to have **lower bias for prediction**, since it uses $y$ to identify directions that are directly relevant for predicting $y$.

**Variance**

- PLS may have **higher variance**, because it uses $y$ to construct components.  
  This makes it more sensitive to noise in $y$, especially in small samples or low signal-to-noise settings.

- PCR often has **lower variance**, since it relies only on the covariance structure of $X$, which is typically more stable than $X^\top y$.

**Summary:**

- PCR: potentially higher bias, lower variance  
- PLS: potentially lower bias, higher variance  

**(b) When PCR Fails**

PCR performs poorly when the directions with large variance in $X$ are not predictive of $y$, while the predictive signal lies in low-variance directions.

A typical construction is:

- $X$ contains:
  - high-variance noise directions
  - low-variance signal directions
- $y$ depends primarily on the low-variance signal

In this case:

- PCA identifies the high-variance directions (noise-dominated)
- PCR keeps these directions and may discard the low-variance signal
- As a result, PCR fails to capture the relationship between $X$ and $y$

**(c) code example**
```{python}
#| eval: false
import numpy as np

rng = np.random.default_rng(0)

n, p = 200, 100

Z_noise = rng.normal(size=(n, 5))
A_noise = rng.normal(size=(5, p))
X_noise = Z_noise @ A_noise * 10

z_signal = rng.normal(size=(n, 1))
a_signal = rng.normal(size=(1, p))
X_signal = z_signal @ a_signal * 0.2

X = X_noise + X_signal +rng.normal(size=(n, p))
y = z_signal.ravel() + 0.5*rng.normal(size=n)

```


:::
:::
























# Week 3: Flexible and Nonparametric Models

## Theory Problems

1. Describe the KNN regression algorithm.

2. Explain how $K$ controls model flexibility.

3. Discuss the bias–variance tradeoff in KNN.

4. Explain the curse of dimensionality.

5. Explain why KNN performs poorly in high dimensions.

6. Compare parametric vs nonparametric regression.

7. Explain how kernel smoothing works.

8. Explain the role of distance metrics in KNN.

9. Explain why larger datasets benefit KNN.

10. Discuss advantages and disadvantages of KNN.

---





# Week 3: KNN, Flexible Regression, and Nonparametric Methods

## Problem 1: KNN Regression Estimator

The KNN regression estimator at $x_0$ is

$$
\hat f(x_0)
=
\frac{1}{K}
\sum_{i \in N_K(x_0)} y_i,
$$

where $N_K(x_0)$ is the set of indices of the $K$ nearest training points.

1. Explain how this estimator works.
2. Explain why $K=1$ gives a very flexible fit.
3. Explain why large $K$ gives a smoother fit.

---

## Problem 2: Bias-Variance in KNN

1. Explain how increasing $K$ changes bias.
2. Explain how increasing $K$ changes variance.
3. Explain why there is usually an optimal value of $K$.

---

## Problem 3: Curse of Dimensionality

1. Explain why nearest neighbors become less meaningful in high dimensions.
2. Explain why nonparametric methods often require more data than parametric methods.

---

## Problem 4: Distance Metrics

1. Explain why standardization matters in KNN.
2. Explain how one predictor with a much larger scale can distort the neighborhood structure.

---

## Problem 5: Comparison with Linear Regression

1. Give an example where linear regression is preferable to KNN.
2. Give an example where KNN is preferable to linear regression.

---

## Problem 6: Simulated Nonlinear Regression

Generate data from

$$
y = \sin(2\pi x) + \varepsilon,
\qquad
\varepsilon \sim N(0, 0.2^2).
$$

1. Fit linear regression.
2. Fit polynomial regression of degree 5.
3. Fit KNN regression for several values of $K$.
4. Compare test MSE.

```{r}
set.seed(432)
# Your code here
```

---

## Problem 7: KNN on the Bike Sharing Data

Use the Bike Sharing dataset.

1. Select a response variable related to rental counts.
2. Split the data into training and testing sets.
3. Fit KNN regression for $K=1,3,5,10,20,50$.
4. Plot test MSE versus $K$.
5. Identify the best $K$.

```{r}
# Your code here
```

---

## Problem 8: Effect of Standardization in KNN

Using the same data:

1. Fit KNN without standardization.
2. Fit KNN with standardization.
3. Compare the results and explain the difference.

```{r}
# Your code here
```

---

## Problem 9: KNN versus Linear Regression

Using the same train/test split:

1. Fit a linear regression model.
2. Compare its test MSE with the best KNN model.
3. Explain which model appears more suitable and why.

```{r}
# Your code here
```

---

## Problem 10: Smoothing Interpretation

Use a one-dimensional predictor/response pair from any open dataset.

1. Fit KNN or another smoother.
2. Plot the fitted curve.
3. Explain how the method captures nonlinearity.

```{r}
# Your code here
```