---
title: "Homework 5"
# author: "Type Your Name Here"
# date: today
jupyter: stats
format: pdf
show_solution: true
---



knn kernel



# Week 5: Ensemble Methods

## Theory Problems

1. Explain bagging.

2. Explain random forests.

3. Explain boosting.

4. Explain variance reduction.

5. Explain feature importance.

6. Compare trees vs forests.

7. Explain bias in boosting.

8. Explain gradient boosting intuition.

9. Explain overfitting in ensembles.

10. Explain out-of-bag error.





# Week 5: Bagging, Random Forests, and Boosting

## Problem 1: Bagging and Variance Reduction

1. Explain why a single tree often has high variance.
2. Explain how bagging reduces variance.
3. Explain why bagging does not usually reduce bias very much.

---

## Problem 2: Random Forests

1. Explain how random forests modify bagging.
2. Explain why selecting only a random subset of predictors at each split helps.
3. Explain what out-of-bag error is.

---

## Problem 3: Boosting

1. Explain the basic idea of boosting.
2. Explain why later trees focus on difficult observations.
3. Explain how boosting can reduce bias.

---

## Problem 4: Feature Importance

1. Explain how variable importance is measured in random forests.
2. Explain why importance should be interpreted carefully.

---

## Problem 5: Trees versus Ensembles

Compare a single tree and a random forest in terms of:

1. interpretability,
2. variance,
3. prediction accuracy.

---

## Problem 6: Titanic Data — Single Tree

Use the Titanic dataset.

1. Fit a classification tree predicting survival.
2. Report training accuracy and test accuracy.
3. Plot the tree.
4. Briefly interpret the first two splits.

```{r}
# Your code here
```

---

## Problem 7: Titanic Data — Random Forest

Using the same train/test split:

1. Fit a random forest.
2. Report test accuracy.
3. Report out-of-bag error.
4. Plot variable importance.

```{r}
# Your code here
```

---

## Problem 8: Random Forest Tuning

1. Fit random forests with different values of `mtry`.
2. Compare out-of-bag error across settings.
3. Identify the best choice of `mtry`.

```{r}
# Your code here
```

---

## Problem 9: Boosting

Using the same dataset or another open classification dataset:

1. Fit a boosting model.
2. Tune the number of trees.
3. Report test accuracy.
4. Compare with the tree and the random forest.

```{r}
# Your code here
```

---

## Problem 10: Ensemble Comparison

Summarize the performance of:

- a single tree,
- bagging or random forest,
- boosting.

Discuss which method performs best and why.

```{r}
# Your code here
```
