---
title: "Template"
# author: "Type Your Name Here"
# date: today
show_solution: false
format: pdf
---





# Problems

::: {#exr-1}
## 1 point

A variable that takes on the values of 0 or 1 and is used to incorporate the effect of categorical variables in a regression model is called
\mc{
\item a constant variable  
\item an interaction   
\item a dummy variable     
\item None of these alternatives is correct.
}
:::


::: {#exr-2}

\mcc[2]{
\item a constant variable  
\item an interaction   
\item a dummy variable     
\item None of these alternatives is correct.
}

:::

```{r}
#| echo: false
dat <- data.frame(x=c(1,2,3), y=c(3,4,3))
```


```{r}
#| echo: false
knitr::kable(
    dat,
    col.names = c("XXX $x$", "YYY $y$"),
     align = "cc"
)
```




::: {.answer}
1. $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2+\beta_4x_1^2+\beta_5x_2^2+\varepsilon$
2. $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2+\varepsilon$
:::



::: {.content-hidden when-meta="show_solution"}
{{< pagebreak >}} 
:::



```{r}
#| echo: true
#| output: false
fit <- lm(y~x, data=dat)
s <- capture.output(summary(fit))
```


```{r}
#| echo: false
#| output: asis

# if (params$show_solution) {
#   s_out <- s        #
# } else {
s_out <- gsub(
    "Multiple R-squared:\\s*[e0-9\\.+-]+",
    "Multiple R-squared:  \\\\red{1. _______}",
    s
)
# }

cat("\\begin{alltt}\n")
cat(paste(s_out, collapse = "\n"))
cat("\n\\end{alltt}\n")
```



```{r}
my_panel <- function(x, y, ...) {
    panel.smooth(x, y, col.smooth = "#BC4B00", ...)
}
plot(fit, which = 5, panel = my_panel, cook.col = "black", cook.lty = 2)
```

