



## Coding Problems

Datasets:
- Mall Customers
- Wholesale Customers

1. Apply K-means clustering.

2. Choose optimal $K$ using elbow method.

3. Visualize clusters.

4. Apply PCA.

5. Plot first two principal components.

6. Compare clustering before/after PCA.

7. Apply hierarchical clustering.

8. Visualize dendrogram.

9. Interpret clusters.

10. Write a short summary.




# Optional Challenge Problems

## Challenge 1: Manual Cross-Validation

Write your own function for $K$-fold cross-validation for one of the following:

- ridge regression,
- lasso,
- KNN.

Do not use a built-in CV wrapper.

```{r}
# Your code here
```

## Challenge 2: Simulation Study of the Curse of Dimensionality

Simulate a nonlinear regression setting and compare KNN performance as the number of predictors increases from $p=1$ to $p=20$.

```{r}
# Your code here
```

## Challenge 3: Final Mini-Project

Choose one open dataset and compare at least three methods from the course. Include:

- train/test split,
- model tuning,
- performance comparison,
- interpretation.

```{r}
# Your code here
```