---
title: "Homework 2"
# author: "Type Your Name Here"
# date: today
jupyter: stats
format: html
show_solution: true
---



# Week 2: Regularization Methods

## Theory Problems



::: {#exr-}
Suppose we estimate the regression coefficients in a linear regression model by minimizing
$$
\sum_{i=1}^n\qty(y_i-\beta_0-\sum_{j=1}^p\beta_jx_{ij})^2+\lambda\sum_{j=1}^p\beta_j^2
$$
for a particular value of $\lambda$. Answer the following questions and justify your answer.

a. As we increase $\lambda$ from $0$, how will the training $\rss$ be changed?
b. As we increase $\lambda$ from $0$, how will the test $\rss$ be changed?
c. As we increase $\lambda$ from $0$, how will the variance be changed?
d. As we increase $\lambda$ from $0$, how will the squared bias be changed?
e. As we increase $\lambda$ from $0$, how will the irreducible error be changed?
:::

::: {.content-visible when-meta="show_solution"}

::: {.solution}

This is ridge regression. The tuning parameter $\lambda$ controls the strength of the penalty:

$$
\lambda \sum_{j=1}^p \beta_j^2.
$$

When $\lambda=0$, the method is ordinary least squares. As $\lambda$ increases, the penalty becomes stronger, so the coefficients are shrunk more toward zero.

**a. Training RSS**

As $\lambda$ increases from $0$, the training RSS will increase or stay the same.

Reason: $\lambda=0$ gives the ordinary least squares solution, which minimizes the training RSS without penalty. Adding a penalty restricts the solution away from the unconstrained RSS minimizer, so the training RSS cannot improve.

**b. Test RSS**

As $\lambda$ increases from $0$, the test RSS will typically decrease at first and then increase.

Reason: a small amount of shrinkage can reduce overfitting and lower variance, improving test performance. But if $\lambda$ becomes too large, the model becomes too constrained, leading to underfitting and larger bias.

**c. Variance**

As $\lambda$ increases from $0$, variance will decrease.

Reason: larger $\lambda$ shrinks the coefficients more strongly toward zero, making the fitted model less sensitive to random variation in the training data.

**d. Squared bias**

As $\lambda$ increases from $0$, squared bias will increase.

Reason: stronger shrinkage pulls the fitted model away from the least squares fit and may prevent it from capturing the true relationship fully.

**e. Irreducible error**

The irreducible error does not change.

Reason: irreducible error is caused by the noise in the data-generating process, not by the choice of $\lambda$ or the regression method.
:::
:::



::: {#exr-}
Suppose we estimate the regression coefficients in a linear regression model by minimizing
$$
\sum_{i=1}^n\qty(y_i-\beta_0-\sum_{j=1}^p\beta_jx_{ij})^2\quad \text{ subject to }\sum_{j=1}^p\abs{\beta_j}\leq s
$$
for a particular value of $s$. Answer the following questions and justify your answer.

a. As we increase $s$ from $0$, how will the training $\rss$ be changed?
b. As we increase $s$ from $0$, how will the test $\rss$ be changed?
c. As we increase $s$ from $0$, how will the variance be changed?
d. As we increase $s$ from $0$, how will the squared bias be changed?
e. As we increase $s$ from $0$, how will the irreducible error be changed?
:::

::: {.content-visible when-meta="show_solution"}

::: {.solution}


This is the lasso constraint form. The parameter $s$ controls how large the coefficients are allowed to be:

$$
\sum_{j=1}^p |\beta_j| \le s.
$$

When $s=0$, all slope coefficients must be zero, so the model is the intercept-only model. As $s$ increases, the constraint becomes less restrictive, and the model becomes more flexible.

**a. Training RSS**

As $s$ increases, the training RSS will decrease or stay the same.

Reason: increasing $s$ enlarges the set of allowable coefficient values. Since the old solution is still feasible, the minimized training RSS cannot increase.

**b. Test RSS**

As $s$ increases, the test RSS will typically decrease at first and then increase.

Reason: when $s$ is very small, the model is too simple and has high bias, so test error is large. Increasing $s$ allows the model to capture more signal, reducing test error. However, if $s$ becomes too large, the model may overfit the training data, increasing variance and therefore increasing test error.

**c. Variance**

As $s$ increases, variance will increase.

Reason: larger $s$ allows more flexible models with larger coefficients. The fitted model becomes more sensitive to the particular training sample, so its variance increases.

**d. Squared bias**

As $s$ increases, squared bias will decrease.

Reason: when $s$ is small, the model is heavily constrained and may underfit the true relationship. Increasing $s$ relaxes the constraint, allowing the model to better approximate the true regression function.

**e. Irreducible error**

The irreducible error does not change.

Reason: irreducible error comes from the random noise in the data-generating process, not from the choice of model or the value of $s$.
:::
:::








::: {#exm-}

## PCR vs PLS

Consider a regression problem with centered data matrix $X \in \mathbb{R}^{n \times p}$ and response $y \in \mathbb{R}^n$.

**a. Bias–Variance Tradeoff**

Discuss how PCR and PLS differ in terms of bias and variance.

- When might PCR have higher bias than PLS?
- When might PLS have higher variance than PCR?


**b. When PCR Fails**

Construct (or describe) a situation where PCR performs way worse than PLS. Explain why PCR may perform poorly in this setting.
<!-- 
**c. When PLS Fails**

Construct (or describe) a situation where PLS performs way worse than PCR. Explain why PLS may perform poorly in this setting. -->

**c. Code examples**

Write Python code to generate the datasets to implement the above two cases. A dataset is a pair $(X,y)$ with both are `numpy.array`.

Ensure there is a performance difference. You can get a bonus if there is at least 1% difference in $R^2$. You may fix random seeds or adjust noise levels to achieve this.

You code should look like this.

```{python}
#| eval: false
import numpy as np

rng = np.random.default_rng(0)

## Write your data generating code here to replace the code below
n, p = 100, 100
X = rng.normal(size=(n, p))
y = rng.normal(size=n)
```

I will test the dataset using the following setting:

1. Split the dataset into trainning and test set with `test_size=0.3`, `random_state=0`.
2. Use `cv=5` and `neg_mean_squared_error` to pick `n_components`. You may set the maximal search cap to be `20` for simplicity.
3. Train the model on the training set, and test the model on the test set.
4. For performance I only look at `r2_score`.


```{python}
#| eval: false
#| echo: false
import numpy as np
from sklearn.model_selection import train_test_split, GridSearchCV
from sklearn.cross_decomposition import PLSRegression
from sklearn.decomposition import PCA
from sklearn.linear_model import LinearRegression
from sklearn.pipeline import Pipeline
from sklearn.preprocessing import StandardScaler
from sklearn.metrics import mean_squared_error, r2_score


def compare_pcr_pls(X, y, max_components=20):

    X_tr, X_te, y_tr, y_te = train_test_split(X, y, test_size=0.3, random_state=0)

    max_components = min(max_components, X_tr.shape[0], X_tr.shape[1])
    param_range = list(range(1, max_components + 1))

    pcr_pipe = Pipeline(
        [
            ("scaler", StandardScaler()),
            ("pca", PCA()),
            ("ols", LinearRegression()),
        ]
    )

    pcr_grid = GridSearchCV(
        pcr_pipe,
        {"pca__n_components": param_range},
        cv=5,
        scoring="neg_mean_squared_error",
    )

    pls_pipe = Pipeline(
        [
            ("scaler", StandardScaler()),
            ("pls", PLSRegression()),
        ]
    )

    pls_grid = GridSearchCV(
        pls_pipe,
        {"pls__n_components": param_range},
        cv=5,
        scoring="neg_mean_squared_error",
    )

    pcr_grid.fit(X_tr, y_tr)
    pls_grid.fit(X_tr, y_tr)

    pcr_mse = mean_squared_error(y_te, pcr_grid.predict(X_te))
    pls_mse = mean_squared_error(y_te, pls_grid.predict(X_te))
    pcr_r2 = r2_score(y_te, pcr_grid.predict(X_te))
    pls_r2 = r2_score(y_te, pls_grid.predict(X_te))

    print(f"PCR test MSE: {pcr_mse}, PCR test R2: {pcr_r2}")
    print(f"PLS test MSE: {pls_mse}, PLS test R2: {pls_r2}")


compare_pcr_pls(X, y)

```


:::

::: {.content-visible when-meta="show_solution"}

::: {.solution}
 **(a) Bias–Variance Tradeoff**

PCR and PLS differ in how they construct components, which leads to different bias–variance behavior.

- **PCR** selects components based only on the variance of $X$, ignoring $y$.
- **PLS** selects components based on their covariance with $y$.

**Bias**

- PCR may have **higher bias** when the directions that are most predictive of $y$ correspond to **low-variance directions in $X$**.  
  In this case, PCR may discard these directions when truncating components.

- PLS tends to have **lower bias for prediction**, since it uses $y$ to identify directions that are directly relevant for predicting $y$.

**Variance**

- PLS may have **higher variance**, because it uses $y$ to construct components.  
  This makes it more sensitive to noise in $y$, especially in small samples or low signal-to-noise settings.

- PCR often has **lower variance**, since it relies only on the covariance structure of $X$, which is typically more stable than $X^\top y$.

**Summary:**

- PCR: potentially higher bias, lower variance  
- PLS: potentially lower bias, higher variance  

**(b) When PCR Fails**

PCR performs poorly when the directions with large variance in $X$ are not predictive of $y$, while the predictive signal lies in low-variance directions.

A typical construction is:

- $X$ contains:
  - high-variance noise directions
  - low-variance signal directions
- $y$ depends primarily on the low-variance signal

In this case:

- PCA identifies the high-variance directions (noise-dominated)
- PCR keeps these directions and may discard the low-variance signal
- As a result, PCR fails to capture the relationship between $X$ and $y$

**(c) code example**
```{python}
#| eval: false
import numpy as np

rng = np.random.default_rng(0)

n, p = 200, 100

Z_noise = rng.normal(size=(n, 5))
A_noise = rng.normal(size=(5, p))
X_noise = Z_noise @ A_noise * 10

z_signal = rng.normal(size=(n, 1))
a_signal = rng.normal(size=(1, p))
X_signal = z_signal @ a_signal * 0.2

X = X_noise + X_signal +rng.normal(size=(n, p))
y = z_signal.ravel() + 0.5*rng.normal(size=n)

```


:::
:::





::: {#exm-}
## Exercise: Ridge vs Lasso

Consider a regression problem with centered data matrix $X \in \mathbb{R}^{n \times p}$ and response $y \in \mathbb{R}^n$.

---

### (a) Bias–Variance Tradeoff

Discuss how ridge regression and lasso differ in terms of bias and variance.

- When might lasso have higher bias than ridge?
- When might ridge have higher variance than lasso?

---

### (b) When Ridge Fails

Construct (or describe) a situation where ridge regression performs much worse than lasso.

Explain your reasoning.

Hint: Consider the case where only a small number of predictors are truly relevant.

---

### (c) Code Examples

Write Python code to generate datasets $(X, y)$ illustrating:

- a case where ridge performs much worse than lasso
- a case where lasso performs worse than ridge

Use cross-validation to tune the regularization parameter.

Ensure there is a performance difference. You can get a bonus if there is at least 1% difference in $R^2$. You may fix random seeds or adjust noise levels to achieve this.

Your code should look like this:

```{python}
#| eval: false
import numpy as np

rng = np.random.default_rng(0)

## Write your data generating code here to replace the code below
n, p = 100, 100
X = rng.normal(size=(n, p))
y = rng.normal(size=n)
```

I will test the dataset using the following setting:

1. Split the dataset into training and test set with `test_size=0.3`, `random_state=0`.
2. Use `cv=5` and `neg_mean_squared_error` to pick $\lambda$.
3. The range for `alpha` is `alphas = np.logspace(-4, 3, 50)`.
4. Train the model on the training set, and test the model on the test set.
5. For performance I only look at `r2_score`.


```{python}
#| eval: false
#| echo: false

import numpy as np
from sklearn.model_selection import train_test_split, GridSearchCV
from sklearn.linear_model import Ridge, Lasso
from sklearn.pipeline import Pipeline
from sklearn.preprocessing import StandardScaler
from sklearn.metrics import mean_squared_error, r2_score


def compare_ridge_lasso(X, y, alphas=None):

    if alphas is None:
        alphas = np.logspace(-4, 3, 50)

    X_tr, X_te, y_tr, y_te = train_test_split(
        X, y, test_size=0.3, random_state=0
    )

    ridge_pipe = Pipeline(
        [
            ("scaler", StandardScaler()),
            ("ridge", Ridge()),
        ]
    )

    ridge_grid = GridSearchCV(
        ridge_pipe,
        {"ridge__alpha": alphas},
        cv=5,
        scoring="neg_mean_squared_error",
    )

    lasso_pipe = Pipeline(
        [
            ("scaler", StandardScaler()),
            ("lasso", Lasso(max_iter=10000)),
        ]
    )

    lasso_grid = GridSearchCV(
        lasso_pipe,
        {"lasso__alpha": alphas},
        cv=5,
        scoring="neg_mean_squared_error",
    )

    ridge_grid.fit(X_tr, y_tr)
    lasso_grid.fit(X_tr, y_tr)

    ridge_pred = ridge_grid.predict(X_te)
    lasso_pred = lasso_grid.predict(X_te)

    ridge_mse = mean_squared_error(y_te, ridge_pred)
    lasso_mse = mean_squared_error(y_te, lasso_pred)

    ridge_r2 = r2_score(y_te, ridge_pred)
    lasso_r2 = r2_score(y_te, lasso_pred)

    print(f"Ridge test MSE: {ridge_mse}, Ridge test R2: {ridge_r2}")
    print(f"Lasso test MSE: {lasso_mse}, Lasso test R2: {lasso_r2}")
```


:::




::: {.content-visible when-meta="show_solution"}

::: {.solution}
**(a) Bias–Variance Tradeoff**

Ridge and lasso differ in how they regularize the model:

- Ridge uses an $\ell_2$ penalty and shrinks all coefficients continuously.
- Lasso uses an $\ell_1$ penalty and can set some coefficients exactly to zero.

Bias

- Lasso may have higher bias when the true signal is dense (many small nonzero coefficients), since it tends to eliminate weaker predictors.
- Ridge generally has lower bias in dense settings, as it retains all predictors.

Variance

- Ridge tends to have lower variance, especially when predictors are highly correlated, since it stabilizes all coefficients.
- Lasso may have higher variance, because variable selection introduces instability (small changes in data can change which variables are selected).

Summary

- Ridge: lower variance, better for dense signals
- Lasso: higher bias in dense settings, better for sparse signals

**(b) When Ridge Fails**

Ridge performs poorly when the true model is sparse:

- Only a small number of predictors are relevant
- Many predictors are pure noise

In this case:

- Ridge keeps all variables with small coefficients
- Noise variables remain and increase prediction error

In contrast:

- Lasso sets many coefficients to zero
- Removes noise variables and improves prediction

Therefore, ridge fails when sparsity is important, while lasso succeeds by performing variable selection.

**(c) Code example**


```{python}
#| eval: false
rng = np.random.default_rng(0)

n, p = 120, 100
X = rng.normal(size=(n, p))

beta = np.zeros(p)
beta[:5] = [10, 8, 6, 4, 2]

y = X @ beta + rng.normal(size=n)
```


```{python}
#| eval: false
#| warning: false

import numpy as np

rng = np.random.default_rng(4)
n = 200
p = 500  
k = 50  

Z = rng.normal(size=(n, k))
A = rng.normal(size=(k, p))
X = Z @ A + 0.01 * rng.normal(size=(n, p))  

gamma = rng.normal(1, 0.1, size=k)
y = Z @ gamma + 3 * rng.normal(size=n)
```



:::
:::



























# Week 2: Ridge Regression, Lasso, and Regularization

## Problem 1: Ridge Regression Formula

Ridge regression solves

$$
\hat{\boldsymbol\beta}^{ridge}
=
\arg\min_{\boldsymbol\beta}
\left\{
\|\mathbf y-\mathbf X\boldsymbol\beta\|_2^2
+
n\lambda \|\boldsymbol\beta\|_2^2
\right\}.
$$

1. Show that the ridge solution is

   $$
   \hat{\boldsymbol\beta}^{ridge}
   =
   (\mathbf X^\top \mathbf X + n\lambda \mathbf I)^{-1}\mathbf X^\top \mathbf y.
   $$

2. Explain why this always exists when $\lambda > 0$.

---

## Problem 2: Ridge Variance Reduction

Assume the linear model from Week 1.

1. Explain why adding $n\lambda \mathbf I$ increases the eigenvalues of $\mathbf X^\top \mathbf X$.
2. Explain why this reduces estimator variance.
3. Explain why this introduces bias.

---

## Problem 3: Bias-Variance Trade-off in Ridge

1. Explain what happens when $\lambda = 0$.
2. Explain what happens as $\lambda \to \infty$.
3. Explain why an intermediate value of $\lambda$ may improve prediction.

---

## Problem 4: Geometric Interpretation

1. Write the constrained form of ridge regression.
2. Explain the geometry of the $L_2$ ball.
3. Explain how this geometry leads to shrinkage but usually not exact zeros.

---

## Problem 5: Lasso Geometry

The lasso solves

$$
\hat{\boldsymbol\beta}^{lasso}
=
\arg\min_{\boldsymbol\beta}
\left\{
\|\mathbf y-\mathbf X\boldsymbol\beta\|_2^2
+
n\lambda \|\boldsymbol\beta\|_1
\right\}.
$$

1. Write the constrained form of lasso.
2. Explain the geometry of the $L_1$ ball.
3. Explain why lasso can set coefficients exactly to zero.

---

## Problem 6: Standardization

1. Explain why predictors should be standardized before applying ridge or lasso.
2. Explain what could go wrong if they are not standardized.

---

## Problem 7: Manual Ridge Prediction

Use the first 24 observations of `mtcars` as training data and the remaining observations as testing data.

1. Fit a ridge regression model with $\lambda = 3$.
2. Compute predictions on the test data manually using

   $$
   \hat y = \mathbf x^\top \hat{\boldsymbol\beta}.
   $$

3. Report the test MSE.

```{r}
library(MASS)
traindata <- mtcars[1:24, ]
testdata <- mtcars[-(1:24), ]
# Your code here
```

---

## Problem 8: Coefficient Paths in Ridge

Using `mtcars` or `Hitters`:

1. Fit ridge regression for a grid of $\lambda$ values.
2. Plot the coefficient paths.
3. Explain which coefficients shrink fastest and why that may happen.

```{r}
library(MASS)
# Your code here
```

---

## Problem 9: Cross-Validation for Ridge

Using `Hitters` from `ISLR2`:

1. Remove missing values.
2. Use 10-fold cross-validation to select $\lambda$ for ridge regression.
3. Report $\lambda_{min}$.
4. Report the corresponding CV error.



---

## Problem 10: Ridge versus Lasso on Real Data

Using the same dataset:

1. Fit ridge regression and lasso with cross-validation.
2. Compare their test errors using a train/test split.
3. For lasso, report the selected variables.
4. Write a brief discussion comparing prediction and interpretability.

```{r}
set.seed(432)
# Your code here
```




ISlR 8
9
