




# Week 2: Regularization Methods

## Theory Problems



::: {#exr-}
## asdf
dsf
:::


1. Explain why multicollinearity leads to unstable OLS estimates.

2. Write the ridge regression optimization problem

$$
\min_\beta \{ RSS + \lambda \sum_{j=1}^p \beta_j^2 \}
$$

Explain how $\lambda$ affects coefficient shrinkage.

3. Explain why ridge regression reduces variance but increases bias.

4. Explain why the lasso penalty

$$
\sum_{j=1}^p |\beta_j|
$$

can produce sparse solutions.

5. Compare ridge regression and lasso in terms of interpretability.

6. Explain why predictors should be standardized before applying ridge or lasso.

7. Describe how cross-validation selects $\lambda$.

8. Explain the geometric difference between $L_1$ and $L_2$ penalties.

9. Suppose $\lambda=0$. What model is obtained?

10. Explain why lasso is useful when $p > n$.

---









# Week 2: Ridge Regression, Lasso, and Regularization

## Problem 1: Ridge Regression Formula

Ridge regression solves

$$
\hat{\boldsymbol\beta}^{ridge}
=
\arg\min_{\boldsymbol\beta}
\left\{
\|\mathbf y-\mathbf X\boldsymbol\beta\|_2^2
+
n\lambda \|\boldsymbol\beta\|_2^2
\right\}.
$$

1. Show that the ridge solution is

   $$
   \hat{\boldsymbol\beta}^{ridge}
   =
   (\mathbf X^\top \mathbf X + n\lambda \mathbf I)^{-1}\mathbf X^\top \mathbf y.
   $$

2. Explain why this always exists when $\lambda > 0$.

---

## Problem 2: Ridge Variance Reduction

Assume the linear model from Week 1.

1. Explain why adding $n\lambda \mathbf I$ increases the eigenvalues of $\mathbf X^\top \mathbf X$.
2. Explain why this reduces estimator variance.
3. Explain why this introduces bias.

---

## Problem 3: Bias-Variance Trade-off in Ridge

1. Explain what happens when $\lambda = 0$.
2. Explain what happens as $\lambda \to \infty$.
3. Explain why an intermediate value of $\lambda$ may improve prediction.

---

## Problem 4: Geometric Interpretation

1. Write the constrained form of ridge regression.
2. Explain the geometry of the $L_2$ ball.
3. Explain how this geometry leads to shrinkage but usually not exact zeros.

---

## Problem 5: Lasso Geometry

The lasso solves

$$
\hat{\boldsymbol\beta}^{lasso}
=
\arg\min_{\boldsymbol\beta}
\left\{
\|\mathbf y-\mathbf X\boldsymbol\beta\|_2^2
+
n\lambda \|\boldsymbol\beta\|_1
\right\}.
$$

1. Write the constrained form of lasso.
2. Explain the geometry of the $L_1$ ball.
3. Explain why lasso can set coefficients exactly to zero.

---

## Problem 6: Standardization

1. Explain why predictors should be standardized before applying ridge or lasso.
2. Explain what could go wrong if they are not standardized.

---

## Problem 7: Manual Ridge Prediction

Use the first 24 observations of `mtcars` as training data and the remaining observations as testing data.

1. Fit a ridge regression model with $\lambda = 3$.
2. Compute predictions on the test data manually using

   $$
   \hat y = \mathbf x^\top \hat{\boldsymbol\beta}.
   $$

3. Report the test MSE.

```{r}
library(MASS)
traindata <- mtcars[1:24, ]
testdata <- mtcars[-(1:24), ]
# Your code here
```

---

## Problem 8: Coefficient Paths in Ridge

Using `mtcars` or `Hitters`:

1. Fit ridge regression for a grid of $\lambda$ values.
2. Plot the coefficient paths.
3. Explain which coefficients shrink fastest and why that may happen.

```{r}
library(MASS)
# Your code here
```

---

## Problem 9: Cross-Validation for Ridge

Using `Hitters` from `ISLR2`:

1. Remove missing values.
2. Use 10-fold cross-validation to select $\lambda$ for ridge regression.
3. Report $\lambda_{min}$.
4. Report the corresponding CV error.



---

## Problem 10: Ridge versus Lasso on Real Data

Using the same dataset:

1. Fit ridge regression and lasso with cross-validation.
2. Compare their test errors using a train/test split.
3. For lasso, report the selected variables.
4. Write a brief discussion comparing prediction and interpretability.

```{r}
set.seed(432)
# Your code here
```
