---
title: "Exam 2"
# author: "Type Your Name Here"
# date: today
format: 
    pdf:        
        pdf-engine: lualatex
        template-partials:
            - ../../doc-class.tex
        include-in-header:
            - ../../math.tex
            - text: |
                \usepackage{xcolor} 
                \usepackage{alltt}
                \usepackage{enumitem}
                \usepackage[letterpaper,top=1in,bottom=0.89in,left=1in,right=1in]{geometry}
                \AtBeginDocument{
                    \renewcommand{\theexercise}{\arabic{exercise}}
                }
                \newcommand{\red}[1]{\textcolor{red}{\textbf{#1}}}
                \newcommand{\mc}[1]{%
                    \begin{enumerate}
                    \setcounter{enumi}{0}
                    \renewcommand{\labelenumi}{\Alph{enumi}.}
                    #1
                    \end{enumerate}
                }
        mainfont: Libertinus Serif
        sansfont: Libertinus Sans
        monofont: Latin Modern Mono
        monofontoptions:
            - Scale=0.95
        mathfont: Libertinus Math
        highlight-style: arrow
params:
    show_solution: false
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---


# Problems

::: {#exr-1}
## 1 point

A variable that takes on the values of 0 or 1 and is used to incorporate the effect of categorical variables in a regression model is called
\mc{
\item a constant variable  
\item an interaction   
\item a dummy variable     
\item None of these alternatives is correct.
}
:::


::: {#exr-2}
## 1 point

If a categorical variable has $k$ levels, _____ dummy variables are required with each dummy variable being coded as 0 or 1.

\mc{
\item $k$  
\item $k-1$    
\item $k-2$    
\item None of these alternatives is correct.   
}
:::


::: {#exr-3}
## 1 point

The following regression model is known as
$$
y=\beta_0+\beta_1x_1+\beta_2x_1^2+\beta_3x_1^3+\varepsilon
$$

\mc{
\item First order model with three predictors  
\item First order model with one predictor     
\item Third order model with one predictor     
\item Third order model with three predictors      
} 

:::


::: {#exr-4}
## 1 point

A variable $x_1x_2$ is added to a multiple regression model in order to account for potential effects of two variables $x_1$ and $x_2$ acting together. This type of effect is

\mc{
\item called multicollinearity effect  
\item called transformation effect     
\item called interaction       
\item called quadratic term        
}
:::


::: {#exr-5}
## 1 point

Which of the following tests is used to determine whether an additional set of several independent variables makes a significant reduction in SSE to a multiple regression model?

\mc{
\item Z-test   
\item t-test   
\item Overall F-test   
\item Partial F-test   
}
:::



::: {.answer}
1. C. a dummy variable  
2. B. $k-1$
3. C. Third order model with one predictor
4. C. called interaction
5. D. Partial F-test
:::

---

**Stroke study** 

A 10-year study conducted relates the probability of a person having stroke over the next 10-year period to person’s age, blood pressure, and whether the person is a smoker as follows:
$$ 
\text{RISK}=−91.8 + 1.08 \text{AGE}+ 0.252 \text{PRESSURE}+ 8.74 \text{SMOKING}
$$
 
where SMOKING is a dummy variable with 1 indicating a smoker and 0 indicating a nonsmoker.


::: {#exr-6}
## 1 point
What is the probability of a stroke over the next 10 years for Art Speen, a 68-year-old smoker who has blood pressure of 175?

\mc{
\item 45.28\%   
\item 34.48\%   
\item 25.74\%   
\item 23.68\%   
}

:::






::: {#exr-7}
## 1 point

Give a practical interpretation of $\hat\beta_1$, the estimated slope of AGE.
:::


::: {.answer}
6. B. 34.48\%  
7. 
:::



---



**Earnings of Mexican street vendors.**

 Detailed interviews were conducted with over 1000 street vendors in the city of Puebla, Mexico, inorder to study the factors influencing vendors’ incomes (World Development, February 1998). Vendorswere defined as individuals working in the street, and included vendors with carts and stands on wheelsand excluded beggars, drug dealers, and prostitutes. The researchers collected data on gender, age, hours worked per day, annual earnings, and education level. An interaction model is fitted to relate the vendor's mean annual earnings $E(y)$ to age ($x_1$) and hours worked ($x_2$). 
$$
E(y)=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2.
$$

```{r}
#| echo: false
source("../../../anova_alt.R")
```

```{r}
#| echo: false

# setwd("D:/Codes/Notes/regression_26s/hidden/Exam/2/")
```


```{r}
#| echo: false
dat <- read.csv("p8.csv")
attach(dat)
```


```{r}
#| echo: false
knitr::kable(
    dat,
    col.names = c("Vendor Number", "Annual Earnings $y$", "AGE $x_1$", "Hours Worked Per Day $x_2$"),
     align = "cccc"
)
```




```{r}
fit_INT <- lm(Earnings~Age*Hours)
anova_alt(fit_INT)
```

```{r}
summary(fit_INT)
```

---

::: {#exr-8}
## 1 point
Give the least squares prediction equation.
:::


::: {#exr-9}
## 1 point
What is the estimated slope relating annual earnings ($y$) to age ($x_1$) when number of hours worked ($x_2$) is 10? (Round to 3 decimal places)
:::


::: {#exr-10}
## 1 point
Test whether the overall model is statistically useful for predicting $y$ using $\alpha=.05$.What is the null and alternative hypothesis?


\mc{
\item $H_0:\beta_0=\beta_1=\beta_2=\beta_3=0$, $H_a:$ At least one of the $\beta$'s $\ne0$
\item $H_0:\beta_1=\beta_2=\beta_3=0$, $H_a:$ At least one of the $\beta$'s $\ne0$
\item $H_0:\beta_1=\beta_2=0$, $H_a:$ At least one of the $\beta$'s $\ne0$
\item $H_0:\beta_3=0$, $H_a:\beta_3\ne0$
\item $H_0:\beta_3=0$, $H_a:\beta_3>0$
}

:::


::: {#exr-11}
## 1 point
Test whether the overall model is statistically useful for predicting $y$ using $\alpha=.05$. What is your conclusion?
:::


::: {#exr-12}
## 1 point
Find the adjusted-$R^2$ value and interpret it.
:::



::: {.answer}
8. a. 
9. b. 22.972
10. c. B. $H_0:\beta_1=\beta_2=\beta_3=0$, $H_a:$ At least one of the $\beta$'s $\ne0$
11. d. 
12. e. 
:::



---

**Carp diet study**
 
Fisheries Science (February 1995) reported on a study of the variables that affect endogenous nitrogenexcretion (ENE) in carp raised in Japan. Carp were divided into groups of 2–15 fish, each according tobody weight and each group placed in a separate tank. The carp were then fed a proteinfree diet threetimes daily for a period of 20 days. One day after terminating the feeding experiment, the amount of ENE in each tank was measured. The table below gives the mean body weight (in grams) and ENEamount (in milligrams per 100 grams of body weight per day) for each carp group.



```{r}
#| echo: false

dat <- read.csv("p13.csv")
attach(dat)
```

```{r}
#| echo: false
knitr::kable(
    dat,
    col.names = c("TANK", "BODY WEIGHT $x$", "ENE $y$"),
     align = "ccc"
)
```


```{r}
#| echo: false

plot(WEIGHT, ENE)
```


```{r}
fit_quad <- lm(ENE~WEIGHT+I(WEIGHT^2))
summary(fit_quad)
```

```{r}
anova_alt(fit_quad)
```

---


::: {#exr-13}
## 1 point
Based on the scatterplot, do you detect a pattern? Describe the nature of the curvature in the plot.
:::



::: {#exr-14}
## 1 point
Fit a quadratic model to the data (R printout shown below). Give the least squares prediction equation.
:::


::: {#exr-15}
## 1 point
Interpret the estimated slope $\hat\beta_2=0.0002735$.
:::



::: {#exr-16}
## 1 point 
Conduct the test $H_0: \beta_2 = 0$ against $H_a: \beta_2\neq0$ using $\alpha=.10$. Give the conclusion in the words of the problem.
:::



::: {.answer}
13. --
14. --
15. --
16. --
:::



---


**Cooling method for gas turbines**
 
Refer to theJournal of Engineering for Gas Turbines and Power(January 2005) study of a high-pressure inlet fogging method for a gas turbine engine. Consider a model for heat rate (kJ/kW-hr) ofa gas turbine as a function of cycle speed (revolutions per minute) and cycle pressure ratio.


```{r}
#| echo: false
dat <- read.csv("p17.csv")
attach(dat)
```


```{r}
fit_SEC <- lm(HEATRATE ~ RPM + CPRATIO + RPM:CPRATIO + I(RPM^2) + I(CPRATIO^2))
anova_alt(fit_SEC)
```


```{r}
fit_FST <- lm(HEATRATE ~ RPM + CPRATIO + RPM:CPRATIO)
anova_alt(fit_FST)
```

```{r}
anova(fit_FST, fit_SEC)
```


---


::: {#exr-17}
## 1 point

Based on the R printout above, which one is the complete second-order model for $y$?

\mc{
    \item $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2+\varepsilon$
    \item $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_1\beta_2x_1x_2+\beta_1^2x_1^2+\beta_2^2x_2^2+\varepsilon$
    \item $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_1\beta_2x_3+\beta_1^2x_4+\beta_2^2x_5+\varepsilon$
    \item $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2+\beta_4x_1^2+\beta_5x_2^2+\varepsilon$
}


:::





::: {#exr-18}
## 1 point


Based on the R printout above, which one is the reduced interaction model for $y$?

\mc{
    \item $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2+\varepsilon$
    \item $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_1\beta_2x_1x_2+\beta_1^2x_1^2+\beta_2^2x_2^2+\varepsilon$
    \item $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_1\beta_2x_3+\beta_1^2x_4+\beta_2^2x_5+\varepsilon$
    \item $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2+\beta_4x_1^2+\beta_5x_2^2+\varepsilon$
}



:::




::: {#exr-19}
## 1 point

Identify the null and alternative hypotheses for determining whether the curvature terms in the complete second-order model are statistically useful for predicting $y$.

\mc{
    \item $H_0: \beta_0=\beta_1=\beta_2=\beta_3=\beta_4=\beta_5=0$, $H_a:$ at least one of the $\beta$'s $\neq0$.
    \item $H_0: \beta_1=\beta_2=\beta_3=\beta_4=\beta_5=0$, $H_a:$ at least one of the $\beta$'s $\neq0$.
    \item $H_0: \beta_4=\beta_5=0$, $H_a:$ at least one of the $\beta$'s $\neq0$.
    \item $H_0: \beta_4=0$, $H_a: \beta_4\neq0$
    \item $H_0: \beta_5=0$, $H_a: \beta_5\neq0$
}




:::


::: {#exr-20}
## 1 point
Based on the R printouts above (of the Complete Model, Reduced Model, and Comparing Nested Model) for the two models above, identify the value of $SSE_R$.
:::




::: {#exr-21}
## 1 point
Based on the R printouts for the two models above, identify the value of $SSE_C$.
:::




::: {#exr-22}
## 1 point
Based on the R printouts for the two models above, identify the value of $MSE_C$.
:::



::: {#exr-23}
## 1 point
What is the value of the F-test statistics for the test of Problem 19? (Round to 4 decimal places)
:::




::: {#exr-24}
## 1 point
Find the rejection region for the test of Problem 19 using $\alpha=0.10$. If use R, which one below gives the critical value $F_{\alpha}$?



```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 3

par(mgp = c(1.5, 0.5, 0))
df1 <- 3
df2 <- 30
alpha <- 0.10

# critical value
Fa <- qf(1 - alpha, df1, df2)

par(xaxs = "i", yaxs = "i")

x <- seq(0, 6, length = 500)
y <- df(x, df1, df2)


plot(x, y,
    type = "l", lwd = 2,
    xlab = "F", ylab = "f(F)", bty = "n", axes = FALSE, ylim = c(0, max(y) * 1.08)
)

axis(2)
axis(1, pos = 0) # ← THIS forces axis at y = 0


# shade right tail
xx <- seq(Fa, 6, length = 200)
polygon(c(Fa, xx, 6), c(0, df(xx, df1, df2), 0),
    col = "grey80", border = NA
)

# mark Fa and alpha
abline(v = Fa, lty = 2)
text(Fa, -0.1, expression(F[alpha]), xpd = TRUE)
text(4.5, 0.08, expression(alpha))
```





\mc{
    \item \texttt{qf(.10, 2, 61)}, which is 0.1055
    \item \texttt{qf(1-.10, 5, 61)}, which is 1.9441
    \item \texttt{qf(1-.10, 3, 63)}, which is 2.1728
    \item \texttt{qf(1-.10, 2, 61)}, which is 2.3917
}
:::



::: {#exr-25}
## 1 point
Which one is the P-value for the test of Problem 19?
\mc{
    \item 2.7903e-27
    \item 7.7198e-26
    \item 0.0002864
}
:::



::: {#exr-26}
## 1 point
Identify the conclusion of the test comparing the nested models in Problem 19 using $\alpha=0.10$.
\mc{
    \item Since P-value $<\alpha$, we reject $H_0$.
    \item Since P-value $<\alpha$, we fail to reject $H_0$.
    \item Since P-value $>\alpha$, we fail to reject $H_0$.
    \item Since P-value $>\alpha$, we reject $H_0$.
}
:::




::: {#exr-27}
## 1 point
Based on the conclusion in the previous question, Which model do we prefer?
\mc{
    \item the complete model
    \item the reduced model
}
:::




::: {.answer}
17. D. $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2+\beta_4x_1^2+\beta_5x_2^2+\varepsilon$
18. A. $y=\beta_0+\beta_1x_1+\beta_2x_2+\beta_3x_1x_2+\varepsilon$
19. C. $H_0: \beta_4=\beta_5=0$, $H_a:$ at least one of the $\beta$'s $\neq0$.
20. 25310639
21. 19370350
22. 317547
23. 9.3534
24. D. `qf(1-.10, 2, 61)`, which is 2.3917
25. C. 0.0002864
26. A. Since P-value $<\alpha$, we reject $H_0$.
27. A. the complete model
:::


