---
title: "Homework 1"
# author: "Type Your Name Here"
# date: today
format: 
    pdf:        
        pdf-engine: lualatex
        template-partials:
            - ../../doc-class.tex
        include-in-header:
            - ../../math.tex
            - text: |
                \usepackage{xcolor}
                \definecolor{AccessibleRed}{HTML}{BC4B00}
                \newcommand{\red}[1]{\textcolor{AccessibleRed}{\textbf{#1}}}
                \usepackage{alltt}
                \usepackage{enumitem}
                \usepackage[letterpaper,top=1in,bottom=0.89in,left=1in,right=1in]{geometry}
                \AtBeginDocument{
                    \renewcommand{\theexercise}{\arabic{exercise}}
                    \setcounter{exercise}{0}
                }                
                \newcommand{\mc}[1]{
                    \begin{enumerate}
                        \setcounter{enumi}{0}
                        \renewcommand{\labelenumi}{\Alph{enumi}.}
                        \setlength{\itemsep}{0pt}
                        \setlength{\topsep}{2pt}
                        #1
                    \end{enumerate}
                }
                \usepackage{multicol}
                \newcommand{\mcc}[2][2]{
                    \begin{multicols}{#1}
                        \begin{enumerate}
                            \setcounter{enumi}{0}
                            \renewcommand{\labelenumi}{\Alph{enumi}.}
                            \setlength{\itemsep}{0pt}
                            \setlength{\topsep}{2pt}
                            #2
                        \end{enumerate}
                    \end{multicols}
                }
        mainfont: Libertinus Serif
        sansfont: Libertinus Sans
        monofont: Latin Modern Mono
        monofontoptions:
            - Scale=0.95
        mathfont: Libertinus Math
        highlight-style: arrow
params:
    show_solution: false
filters:
    - ../../solution-filter.lua
crossref:
    chapters: false
---






---

# Coding Problems

Datasets:
- Boston Housing
- Auto MPG
- California Housing

---

1. Load the **Boston dataset** and fit a linear regression model predicting `medv`.

Compute:

- training $R^2$
- residual diagnostics.

---

2. Split the Boston dataset into training and test sets.

Compute the **test MSE**.

---

3. Using the **Auto dataset**, fit a regression model predicting `mpg`.

Evaluate prediction accuracy.

---

4. Fit polynomial regression models of degree 1–5 predicting `mpg` from `horsepower`.

Plot training error vs model degree.

---

5. Repeat Problem 4 using a **train/test split** and plot test error.

---

6. Fit a regression model predicting house prices using the **California Housing dataset**.

Report $R^2$.

---

7. Plot residuals vs fitted values and interpret the plot.

---

8. Remove one important predictor and compare model performance.

---

9. Fit two competing models and compare **test MSE**.

---

10. Write a short summary describing which model performs best.

---





# Week 1: Foundations, Bias-Variance Trade-off, and Linear Regression

## Problem 1: Bias-Variance Decomposition

Suppose

$$
Y = f(X) + \varepsilon,
$$

where

$$
E(\varepsilon \mid X)=0, \qquad \mathrm{Var}(\varepsilon \mid X)=\sigma^2.
$$

Let $\hat f(x_0)$ be an estimator of $f(x_0)$ based on a training sample, and let

$$
Y_0 = f(x_0) + \varepsilon_0
$$

be a future response at the predictor value $x_0$.

1. Show that the expected prediction error at $x_0$ is

   $$
   E\big[(Y_0-\hat f(x_0))^2\big].
   $$

2. Derive the decomposition

   $$
   E\big[(Y_0-\hat f(x_0))^2\big]
   =
   \mathrm{Var}(\hat f(x_0))
   +
   \big(E[\hat f(x_0)]-f(x_0)\big)^2
   +
   \sigma^2.
   $$

3. Explain the meaning of each term.

---

## Problem 2: OLS Unbiasedness

Consider the linear model

$$
\mathbf y = \mathbf X \boldsymbol\beta + \boldsymbol\varepsilon,
$$

where

$$
E(\boldsymbol\varepsilon)=\mathbf 0, \qquad
\mathrm{Var}(\boldsymbol\varepsilon)=\sigma^2 \mathbf I.
$$

Assume $\mathbf X$ has full column rank.

1. Show that the OLS estimator is

   $$
   \hat{\boldsymbol\beta}
   =
   (\mathbf X^\top \mathbf X)^{-1}\mathbf X^\top \mathbf y.
   $$

2. Prove that $\hat{\boldsymbol\beta}$ is unbiased.
3. Show that

   $$
   \mathrm{Var}(\hat{\boldsymbol\beta})
   =
   \sigma^2(\mathbf X^\top \mathbf X)^{-1}.
   $$

---

## Problem 3: Correlated Predictors and Variance Inflation

Suppose two predictors $x_1$ and $x_2$ are highly correlated.

1. Explain why the matrix $\mathbf X^\top \mathbf X$ becomes nearly singular.
2. Explain why this increases the variance of $\hat{\boldsymbol\beta}$.
3. Explain why coefficient instability does not necessarily imply poor training fit.

---

## Problem 4: Model Complexity

Suppose two models are fitted to the same dataset:

- Model A: simple linear regression
- Model B: polynomial regression of degree 10

Answer the following.

1. Which model must have smaller or equal training RSS?
2. Which model is likely to have higher variance?
3. Under what conditions could Model B have larger test error?

---

## Problem 5: Parametric versus Nonparametric Models

1. Explain the difference between a parametric model and a nonparametric model.
2. Give one example of each.
3. Compare them in terms of flexibility, interpretability, and sample size requirement.

---

## Problem 6: Simulation of OLS under a Correctly Specified Model

Generate data from

$$
y_i = 3 + 2x_i + \varepsilon_i,
\qquad
\varepsilon_i \sim N(0,1),
$$

with $n=100$ and $x_i \sim \mathrm{Unif}(0,1)$.

1. Generate one dataset and fit the model $y \sim x$.
2. Repeat the simulation 500 times.
3. Store the estimated slope and intercept from each repetition.
4. Plot histograms of the estimated slopes and intercepts.
5. Compare the empirical mean of the estimated slope with the true value $2$.

```{r}
set.seed(432)
# Your code here
```

---

## Problem 7: Simulation under Misspecification

Now generate data from

$$
y_i = 3 + 2x_i + x_i^2 + \varepsilon_i,
\qquad
\varepsilon_i \sim N(0,1),
$$

with $n=100$ and $x_i \sim \mathrm{Unif}(0,1)$, but still fit the linear model

$$
y \sim x.
$$

1. Repeat the experiment 500 times.
2. Plot the distribution of the estimated slope.
3. Compare the average fitted slope with the true linear coefficient $2$.
4. Explain the result.

```{r}
set.seed(432)
# Your code here
```

---

## Problem 8: Auto Dataset — Simple Linear Regression

Use the `Auto` dataset from `ISLR2`.

1. Fit the model

   $$
   \texttt{mpg} \sim \texttt{horsepower}.
   $$

2. Report the estimated coefficients, $R^2$, and residual standard error.
3. Plot the data and add the fitted regression line.
4. Comment on whether the linear fit seems adequate.



---

## Problem 9: Polynomial Regression and Test Error

Using the `Auto` dataset:

1. Randomly split the data into training (70%) and testing (30%).
2. Fit polynomial models of degree $1,\ldots,5$ predicting `mpg` from `horsepower`.
3. Compute the training MSE and testing MSE for each degree.
4. Plot degree versus training MSE and degree versus testing MSE.
5. Explain the pattern.

```{r}
set.seed(432)
# Your code here
```

---

## Problem 10: Cross-Validation and Model Selection

Again using the `Auto` dataset:

1. Use 10-fold cross-validation to compare polynomial models of degree $1,\ldots,10$.
2. Identify the degree with the smallest CV error.
3. Compare your answer with Problem 9.
4. Explain why the chosen degree may differ between the test-set approach and cross-validation.

```{r}
set.seed(432)
# Your code here
```
